---
title: "Brownlows Bets"
author: "Yaretzi Padilla"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "Fall 2026"    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(tigerstats)
require(knitr)

```


<!-- We recommend you use git branches -- create a separate branch for each major section i.e. Introduction, Methods, Results, Inferential Results, and Discussion and Conclusions - leave the YAML block untouched until the final merge where you put in your own name !-->

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

 At the first day of class there are 35 students in Math Statistics class and professor Brownlow always proposes this bet to his students "I will wager $2 to your bet of $1 that at least two people in today's class have  the same birthday.  He tells me he has made that bet 637 times so far. Where do you think his profits are? Discuss this in a markdown document
 

There are 3 questions:   
1. What are the fair odds for a bet on matching birthdays in a 35 person class.  
2. For what size class is a 2:1 odds bet fair.  
3. How much money would Brownlow be expected to end up with if the bets went as described.  

Our strategy is to build a Function and the use it to answer the three questions. 

```{r}
# set up the 'urn' with all birthdays
urn = c(1:365)

nStudents = 28 
print(paste('number of students in class:', nStudents))

```
## Question 1
In the class and payoff as described-28 students, payoff 2:1,  what is Brownlow's average over 200 Monte Carlo simulations?  Find the mean amount, standard deviation, and a 95% confidence interval for the earnings.

<!--Explain and display your method of coming up with a 95% confidence interval of  professor Brownlows winnings (or losings)-->

```{r}
set.seed(116)

urn = 1:365
nStudents = 28
nSimulations = 200
results = numeric(nSimulations)

for(i in 1:nSimulations) {
  samplebday = sample(urn, size = nStudents, replace = TRUE)
  l = length(unique(samplebday))

  if(l < nStudents) {
    results[i] = 1
  } else {
    results[i] = -2
  }
}

mean(results)
sd(results)
t.test(results, conf.level = 0.95)
```
After conducting 200 Monte Carlo simulations with 28 students, Brownlow's average earnings were $0.025 per bet. The standard deviation was approximately $1.41, indicating variation in his winnings and losses. Also, the 95% confidence interval ranged from −$0.17142 to $0.22142. Since the interval includes zero, the simulation does not provide sufficient evidence that Brownlow consistently earns or loses money under these betting conditions

## 
What are the fair odds? Use manipulation of odds in the function to tune Brownlow's earnings to zero. 


<!--Decide on your methods:  In this section explain the approach you will take to attack the issue --- how will you decide on a likely estimate of professor Brownlow's earnings or losses after 637 bets. Also explain how you construct a 95% confidence interval about your estimate of his winnings !-->



## 
What size of class makes the 2:1 bet fair?  Use manipulation of number of Students to tune Brownlow's earnings to zero.

<!--Explain and display the results of your method of coming up with the estimate of professor Brownlows betting profits!-->



# Discussion and Conclusion

<!--Discussion and conclusion here.  If professor Brownlow is consistantly making or losing money how could he change his payout so that his profits center on zero?  Also discuss what size his class should be in order to have his current scheme balance at zero profit.!-->

